### 2018-03-16 paper/5-implementations.tex
Mostly TODOs for S5

\clearpage
\section{Build systems, concretely}\label{sec-implementations}

In the previous sections we discussed the types of build system, and how they broken down. But these divisions were not obvious to us, and only by concretely implementing and refactoring each build system did we determine the underlying commonalities. In this section we share some of the code that got us there, but full code is available in the supporting material and at \url{url-hidden-for-anonymous-submission}. Our code comprises a framework, utility functions implementing the various dependency orderings as per \S\ref{sec-dependency-orderings}, and actual implementations.

\subsection{Framework}

todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
there anything that remains?}

As we saw in \S\ref{sec-build}, a build system can be defined as:

\begin{minted}{haskell}
type Build c i k v = Compute c k v -> k -> Maybe i -> (k -> v) -> (i, k -> v)
\end{minted}

Using the framework we build in this section, we can express a dumb build system as:

\begin{minted}{haskell}
dumb :: Build Monad () k v
dumb compute k = runM (f k)
    where
        f k = case compute f k of
            Nothing -> getStore k
            Just act -> do
                v <- act
                putStore k v
                return v
\end{minted}

The \hs{dumb} build system can build with monadic computes, using a function \hs{f} which given a key returns the associated value. To do so, \hs{f} calls \hs{compute} passing \hs{f} itself as an argument. If the result is \hs{Nothing} (this key is an input) it simply reads from the store. Otherwise it runs the action to produce a value, stores it and returns it.

\begin{figure}
\begin{minted}{haskell}
type M i k v r = State (i, k -> v, ...) r
runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)

getStore :: k -> M i k v v
putStore :: k -> v -> M i k v ()
getInfo :: M i k v i
modifyInfo :: (i -> i) -> M i k v ()
\end{minted}
Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
\caption{API with which to implement build systems}
\label{fig-M-api}
\end{figure}

We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}. We provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store. We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.

@@

@@

\begin{minted}{haskell}
make :: Eq v => Build Applicative (Changed k v, ()) k v
make = withChanged $ topological $ \k dk act -> do
    t <- getStoreTime k
    dt <- mapM getStoreTime dk
    let dirty = any (>= t) dt
    when dirty (putStore k =<< act)
\end{minted}

@@

\begin{minted}{haskell}
topological
    :: (m ~ M i k v, Default i)
    => (k -> [k] -> m v -> m ())
    -> Build Applicative i k v
topological step compute k = runM $ do
    let depends = dependencies compute
    forM_ (topSort depends $ transitiveClosure depends k) $ \k ->
        case compute getStore k of
            Nothing -> return ()
            Just act -> step k (depends k) act
\end{minted}

@@

\begin{minted}{haskell}
getChanged :: Eq v => k -> M (Changed k v, i) k v Bool
withChanged :: Build c (Changed k v, i) k v -> Build c (Changed k v, i) k v

data Time = LastBuild | AfterLastBuild
getStoreTime :: Eq v => k -> M (Changed k v, i) k v Time
\end{minted}

To model a dirty bit we extend the \hs{i} information with \hs{Changed} to capture the \hs{k -> v} store at the end of the previous run, and extend \hs{M} to record which files have been written to during this \hs{runM}. A file is considered dirty if either the current value as per \hs{getStore} has changed since the last run, or if it has had \hs{putStore} called on it\footnote{It would be reasonable to declare files unchanged if they have \hs{putStore} called at their previous value, but \Excel does not and \Make cannot using modification times, so we ignore this option.}. To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes\footnote{For tedious reasons our implementation requires \hs{withChangedApplicative} and \hs{withChangedMonad} which are variants of \hs{withChanged} for concrete constraints \hs{c} - but with identical implementations.}

@@

@@

@@

\subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}

>>>>

\section{Build systems, concretely}\label{sec-implementations}

In the previous sections we discussed the types of build system, and how they
can be broken down. But these divisions were not obvious to us, and only by
concretely implementing and refactoring each build system did we determine the
underlying commonalities. In this section we share some of the code that got us
there. Our code comprises a framework, utility functions implementing the
various dependency orderings as per~\S\ref{sec-dependency-orderings}, and
actual implementations.

% \subsection{Framework}

% todo{SLPJ}{I think this entire section, including the mysterious M, can be deleted.  Or is
% there anything that remains?}

% \begin{figure}
% \begin{minted}{haskell}
% type M i k v r = State (i, k -> v, ...) r
% runM :: Default i => M i k v a -> Maybe i -> (k -> v) -> (i, k -> v)

% getStore :: k -> M i k v v
% putStore :: k -> v -> M i k v ()
% getInfo :: M i k v i
% modifyInfo :: (i -> i) -> M i k v ()
% \end{minted}
% Note: we have omitted the context on \hs{k}, things like \hs{Eq} (to implement the store).
% \caption{API with which to implement build systems}
% \label{fig-M-api}
% \end{figure}

% We use \hs{runM} to switch between the pure world of \hs{Build} and a state monad \hs{M} which lets us more precisely express the ideas behind the build system without threading the state manually - the types of important operations are given in Figure~\ref{fig-M-api}. We provide functions like \hs{getStore} to load from the store and \hs{putStore} to write to the store. We'll introduce the other functions in in the monad \hs{M} as we need them while exploring the build systems.

@@

\todo{AM}{Explain the new implementation. Most of the explanations below will
need to be adapted.}

@@

\begin{minted}{haskell}
type Time = Integer
type MakeInfo k = (k -> Time, Time)

make :: Eq k => Build Applicative (MakeInfo k) k v
make = topological $ \key deps act -> do
    (modTime, now) <- getInfo <$> get
    let dirty = or [ modTime dep > modTime key | dep <- deps ]
    when dirty $ do
        v <- act
        let newModTime k = if k == key then now else modTime k
        modify $ \s -> putInfo (putValue s key v) (newModTime, now + 1)
\end{minted}

@@

\begin{minted}{haskell}
topological :: Eq k
            => (k -> [k] -> State (Store i k v) v -> State (Store i k v) ())
            -> Build Applicative i k v
topological step task key = execState $ forM_ chain $ \k -> do
    let fetch k = do { store <- get; return (getValue store k) }
    case task fetch k of
        Nothing  -> return ()
        Just act -> step k (deps k) act
  where
    deps  = dependencies task
    chain = topSort deps (closure deps key)
\end{minted}

@@

To model a dirty bit we extend the \hs{i} information with \hs{Changed} to capture the \hs{k -> v} store at the end of the previous run, and extend \hs{M} to record which files have been written to during this \hs{runM}. A file is considered dirty if either the current value as per \hs{getStore} has changed since the last run, or if it has had \hs{putStore} called on it\footnote{It would be reasonable to declare files unchanged if they have \hs{putStore} called at their previous value, but \Excel does not and \Make cannot using modification times, so we ignore this option.}. To track the values at the last run we use \hs{withChanged} to copy the resultant \hs{k -> v} store into the info after the build completes.

@@

\todo{AM}{I might do some simplification of this tomorrow. Neil: I guess it's best
for you to focus on nailing traces and associated \Shake and \Bazel sections.}

@@

\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}

@@

\subsection{Smarter \hs{[Trace]} Structures}\label{sec-smart-traces}
\todo{NM}{Remove the intermediate \hs{M} layer, as done in \Make.}
